\LARGE{\bf{Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})}}\footnotemark

### 2014-05-20 manuscript.tex
New Abstract

\emph{Premise of the study:} The \emph{teosinte branched1} (\emph{tb1}) gene, a repressor of lateral organ growth, is a major QTL involved in branching differences between maize and its wild progenitor, teosinte. Previous work has shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances the gene's expression, causing much of the reduction in tillering observed in domesticated maize. Observations of the maize \emph{tb1} allele in teosinte individuals, coupled with estimates of an age of insertion of the \emph{Hopscotch} element that predates domestication, led us to investigate its prevalence and potential role in teosinte.

@@

\emph{Methods:}

@@

\emph{Key results:} Results from genotyping across many natural populations suggest that the \emph{Hopscotch} element is segregating at a higher than expected frequency in a number of populations of two subspecies of teosinte, \emph{Zea mays} ssp. \emph{parviglumis} and \emph{Zea mays} ssp. \emph{mexicana}. Analysis of linkage disequilibrium between the \emph{Hopscotch} element and variation in surrounding regions does not support a hypothesis of recent introgression from maize into teosinte.  Population genetic signatures, however, are consistent with selection on this locus and suggest the \emph{Hopscotch} insertion at \emph{tb1} may play an ecological role in teosinte. Finally, two greenhouse experiments with \emph{Zea mays} ssp. \emph{parviglumis} do not suggest \emph{tb1} controls tillering in natural populations of this subspecies.

@@

\emph{Conclusions:} Our findings suggest that the role of the \emph{Hopscotch} in tillering in teosinte differs from domesticated maize, and that other loci may play a role in observed variation in this trait.

@@

>>>>

\emph{Premise of the study:} The \emph{teosinte branched1} (\emph{tb1}) gene is a major QTL controlling branching differences between maize and its wild progenitor, teosinte. Previous work has shown that the insertion of a transposable element (\emph{Hopscotch}) upstream of \emph{tb1} enhances the gene's expression, causing much of the reduction in tillering observed in maize. Observations of the maize \emph{tb1} allele in teosinte and estimates of an age of insertion of the \emph{Hopscotch} element that predates domestication led us to investigate its prevalence and potential role in teosinte.

@@

\emph{Methods:} Prevalence of the \emph{Hopscotch} element was assessed across an Americas-wide sample of 1110 maize and teosinte individuals using a co-dominant PCR assay.  Population genetic summaries were calculated for a subset of individuals from four teosinte populations in central Mexico. Phenotypic data were also collected from a single teosinte population where \emph{Hopscotch} was found segregating.

@@

\emph{Key results:} Genotyping results suggest the \emph{Hopscotch} element is at higher than expected frequency in teosinte. Analysis of linkage disequilibrium near \emph{tb1} does not support recent introgression of the \emph{Hopscotch} allele from maize into teosinte.  Population genetic signatures are consistent with selection on this locus revealing a potential ecological role for \emph{Hopscotch} in teosinte. Finally, two greenhouse experiments with teosinte do not suggest \emph{tb1} controls tillering in natural populations.

@@

\emph{Conclusions:} Our findings suggest the role of \emph{Hopscotch} differs between maize and teosinte. Future work should assess \emph{tb1} expression levels in teosinte with and without the \emph{Hopscotch} and more comprehensively phenotype teosinte to assess the ecological significance of the \emph{Hopscotch} insertion and, more broadly, the \emph{tb1} locus in teosinte.

@@

\textbf{Key words:} domestication; maize; teosinte; \emph{teosinte branched1}
